A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-04-11 content/manuscript.qmd
incorporate minor changes & add acknowledgements

  - name: Yoel Perez Haas

@@

  - name: Bertrand Poumayou

@@

@@

Such a personalized risk prediction may allow clinicians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life [@batth_practical_2014].

Here, we extend the previous work by incorporating the primary tumor location (specified as ICD-10 code) into the model of lymphatic tumor progression, focusing on tumors in the oropharynx and the oral cavity. HNSCC patients with primary tumors at different subsites show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999]. So far, this could be handled by training different models for broader categories of tumor locations, e.g. oropharynx and oral cavity tumors. However, this approach does not describe differences in lymphatic spread between different subsites within the oropharynx and oral cavity. To address this issue, we present an approach using mixtures of HMMs. The intuition is that the lymphatic spread of a tumor that lies anatomically at the boarder of oropharynx and oral cavity (e.g. tumors in the palate) may be described by a mixture of different models.

@@

where $\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}$ denotes the set of all model parameters (7 in our case). Fortunately, the exact length and shape of this distribution has little impact as previously shown [@ludwig_hidden_2021]. We set $t_\text{max}=$ {{< var model.max_t >}} and $P_\text{early}(t)$ to a binomial distribution with parameter {{< var model.first_binom_prob >}}. Further details on the HMM can be found in @ludwig_hidden_2021.

![On the left: Anatomical sketch of the tumor subsites and correspsonding ICD-10 codes considered in this work. The subsite "other parts of mouth" (C06) was not drawn. On the right: Parametrized graphical model of the lymphatic network considering four LNLs. Blue nodes represent the hidden states of LNLs $X_v$, while the red one is the tumor. Arcs represent possible routes of metastatic spread, associated with a probability.](../static/schematic_and_graph.svg){#fig-schematic-and-graph fig-pos="th"}

@@

>>>>

  - name: Yoel Pérez Haas

@@

  - name: Bertrand Pouymayou

@@

    link-citations: true

@@

Such a personalized risk prediction may allow clinicians to safely reduce the elective clinical target volume (CTV-N) and thus reduce side-effects that degrade the patient's quality of life without compromising treatment efficacy [@batth_practical_2014].

Here, we extend the previous work by incorporating the primary tumor location (specified as ICD-10 code) into the model of lymphatic tumor progression, focusing on tumors in the oropharynx and the oral cavity. HNSCC patients with primary tumors at different subsites show different patterns of lymphatic spread [@lindberg_distribution_1972;@woolgar_histological_1999]. So far, this could be handled by training different models for broader categories of tumor locations, e.g. oropharynx and oral cavity tumors. However, this approach does not describe differences in lymphatic spread between different subsites within the oropharynx and oral cavity. To address this issue, we present an approach using mixtures of HMMs. The intuition is that the lymphatic spread of a tumor that lies anatomically at the boarder of oropharynx and oral cavity (e.g. tumors in the palate) may be described by a mixture of different models. Tumor subsites used in this work are sketched in @fig-schematic-and-graph.

@@

where $\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}$ denotes the set of all model parameters (7 in our case). Fortunately, the exact length and shape of this distribution has little impact as previously shown [@ludwig_hidden_2021]. We set $t_\text{max}=$ {{< var model.max_t >}} and $P_\text{early}(t)$ to a binomial distribution with parameter {{< var model.first_binom_prob >}}. Further details on the HMM can be found in @ludwig_hidden_2021 and @zora231470.

![On the left: Anatomical sketch of the tumor subsites and corresponding ICD-10 codes considered in this work. The subsite "other parts of mouth" (C06) was not drawn. On the right: Parametrized graphical model of the lymphatic network considering four LNLs. Blue nodes represent the hidden states of LNLs $X_v$, while the red one is the tumor. Arcs represent possible routes of metastatic spread, associated with a probability.](../static/schematic_and_graph.svg){#fig-schematic-and-graph fig-pos="th"}

@@

# Acknowledgements {.unnumbered}

This work was supported by the clinical research priority program (CRPP) "Artificial intelligence in oncological imaging" of the University of Zurich and by the Swiss cancer research foundation under 
grant KFS 5645–08–2022.

### 2024-04-11 content/manuscript.tex
incorporate minor changes & add acknowledgements

\author[%
1,2%
]{Roman Ludwig}
\author[%
1,2%
]{Julian Brönnimann}
\author[%
1,2%
]{Yoel Perez Haas}
\author[%
1,2%
]{Esmée Lauren Looman}
\author[%
2%
]{Panagiotis Balermpas}
\author[%
11%
]{Sergi Benavente}
\author[%
3,4,7%
]{Adrian Schubert}
\author[%
8%
]{Dorothea Barbatei}
\author[%
8%
]{Laurence Bauwens}
\author[%
2%
]{Jean-Marc Hoffmann}
\author[%
3%
]{Olgun Elicin}
\author[%
6,10%
]{Matthias Dettmer}
\author[%
2%
]{Bertrand Poumayou}
\author[%
4,5%
]{Roland Giger}
\author[%
8%
]{Vincent Grégoire}
\author[%
1,2%
]{Jan Unkelbach}

@@

Such a personalized risk prediction may allow clinicians to safely
reduce the elective clinical target volume (CTV-N) and thus reduce
side-effects that degrade the patient's quality of life
\autocite{batth_practical_2014}.

Here, we extend the previous work by incorporating the primary tumor
location (specified as ICD-10 code) into the model of lymphatic tumor
progression, focusing on tumors in the oropharynx and the oral cavity.
HNSCC patients with primary tumors at different subsites show different
patterns of lymphatic spread
\autocite{lindberg_distribution_1972,woolgar_histological_1999}. So far,
this could be handled by training different models for broader
categories of tumor locations, e.g.~oropharynx and oral cavity tumors.
However, this approach does not describe differences in lymphatic spread
between different subsites within the oropharynx and oral cavity. To
address this issue, we present an approach using mixtures of HMMs. The
intuition is that the lymphatic spread of a tumor that lies anatomically
at the boarder of oropharynx and oral cavity (e.g.~tumors in the palate)
may be described by a mixture of different models.

@@

where \(\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}\) denotes
the set of all model parameters (7 in our case). Fortunately, the exact
length and shape of this distribution has little impact as previously
shown \autocite{ludwig_hidden_2021}. We set \(t_\text{max}=\) 10 and
\(P_\text{early}(t)\) to a binomial distribution with parameter 0.3.
Further details on the HMM can be found in
\textcite{ludwig_hidden_2021}.

@@

\caption{\label{fig-schematic-and-graph}On the left: Anatomical sketch
of the tumor subsites and correspsonding ICD-10 codes considered in this
work. The subsite ``other parts of mouth'' (C06) was not drawn. On the
right: Parametrized graphical model of the lymphatic network considering
four LNLs. Blue nodes represent the hidden states of LNLs \(X_v\), while
the red one is the tumor. Arcs represent possible routes of metastatic
spread, associated with a probability.}

@@

>>>>

\author[%
1,2%
]{Roman Ludwig}
\author[%
1,2%
]{Julian Brönnimann}
\author[%
1,2%
]{Yoel Pérez Haas}
\author[%
1,2%
]{Esmée Lauren Looman}
\author[%
2%
]{Panagiotis Balermpas}
\author[%
11%
]{Sergi Benavente}
\author[%
3,4,7%
]{Adrian Schubert}
\author[%
8%
]{Dorothea Barbatei}
\author[%
8%
]{Laurence Bauwens}
\author[%
2%
]{Jean-Marc Hoffmann}
\author[%
3%
]{Olgun Elicin}
\author[%
6,10%
]{Matthias Dettmer}
\author[%
2%
]{Bertrand Pouymayou}
\author[%
4,5%
]{Roland Giger}
\author[%
8%
]{Vincent Grégoire}
\author[%
1,2%
]{Jan Unkelbach}

@@

Such a personalized risk prediction may allow clinicians to safely
reduce the elective clinical target volume (CTV-N) and thus reduce
side-effects that degrade the patient's quality of life without
compromising treatment efficacy \autocite{batth_practical_2014}.

Here, we extend the previous work by incorporating the primary tumor
location (specified as ICD-10 code) into the model of lymphatic tumor
progression, focusing on tumors in the oropharynx and the oral cavity.
HNSCC patients with primary tumors at different subsites show different
patterns of lymphatic spread
\autocite{lindberg_distribution_1972,woolgar_histological_1999}. So far,
this could be handled by training different models for broader
categories of tumor locations, e.g.~oropharynx and oral cavity tumors.
However, this approach does not describe differences in lymphatic spread
between different subsites within the oropharynx and oral cavity. To
address this issue, we present an approach using mixtures of HMMs. The
intuition is that the lymphatic spread of a tumor that lies anatomically
at the boarder of oropharynx and oral cavity (e.g.~tumors in the palate)
may be described by a mixture of different models. Tumor subsites used
in this work are sketched in figure~\ref{fig-schematic-and-graph}.

@@

where \(\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}\) denotes
the set of all model parameters (7 in our case). Fortunately, the exact
length and shape of this distribution has little impact as previously
shown \autocite{ludwig_hidden_2021}. We set \(t_\text{max}=\) 10 and
\(P_\text{early}(t)\) to a binomial distribution with parameter 0.3.
Further details on the HMM can be found in \textcite{ludwig_hidden_2021}
and \textcite{zora231470}.

@@

\caption{\label{fig-schematic-and-graph}On the left: Anatomical sketch
of the tumor subsites and corresponding ICD-10 codes considered in this
work. The subsite ``other parts of mouth'' (C06) was not drawn. On the
right: Parametrized graphical model of the lymphatic network considering
four LNLs. Blue nodes represent the hidden states of LNLs \(X_v\), while
the red one is the tumor. Arcs represent possible routes of metastatic
spread, associated with a probability.}

@@

\section*{Acknowledgements}\label{acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

This work was supported by the clinical research priority program (CRPP)
``Artificial intelligence in oncological imaging'' of the University of
Zurich and by the Swiss cancer research foundation under grant KFS
5645--08--2022.
